Dynamical Properties of Molecular Cloud Complexes in Galaxies at the Epoch of Reionization

### 2019-03-27 WRITEUP/ms.tex
rearranged Toomre discussion fairly extensively, added reference to Rafikov 2001

The onset of gravitational instability is considered to be tightly connected to \SF \citep[e.g.,][]{Kennicutt89a, Wang94a, Li05b, Li06a}. From virial theorem, Jeans instability sets the scales at which gravitational potential overcomes the internal energy. Namely, gravitational instabilities dominates at scales $L > L_J$, where $L_J$ is the Jeans length. However, centrifugal forces resulting from differential rotation in disk galaxies can stabilize perturbations even on these scales. As a result, a disk is thought to be unstable on scales between $L_J < L < L_{\rm rot}$, where $L_{\rm rot}$ is set by the surface density of the disk and the angular velocity of the rotation.

In theoretical and observational works, gravitational instability has been widely discussed in the context of the Toomre $Q$ stability criterion \citep{Toomre64a, Goldreich65b} or its observable proxy  $v_{\rm circ}/\sigma$, i.e., the circular velocity to the disk r.m.s. velocity dispersion \AF{This is not exactly the same $\sigma$ as before} ratio \citep[e.g.,][]{GarciaBurillo03a, Genzel11a, Kassin12a, Leung19a}.
% https://ned.ipac.caltech.edu/level5/March15/Glazebrook/Glazebrook5.html
%
% from linear analysis
For axisymmetric modes, the dispersion relation for the growth of density perturbations in a rotating, turbulent disk of finite thickness $h$ is described by
\begin{equation}
\omega^2 = \kappa^2 - \frac{2\pi G \Sigma |k|}{1 + |k| h} + \sigma^2 k^2\,,
\label{eqn:3Ddisp}
\end{equation}
where $k$ is the wavenumber and $\kappa$ is the epicyclic frequency, defined as:
\begin{equation}
\kappa^2\equiv\frac{2\Omega}{R}\frac{d}{dR}\left(R^2\Omega\right)\,.
\label{eqn:kappa}
\end{equation}
\citep{Romeo92a}.
\AF{$R$ is not the same as the MCC radius in the proev. Section}
In Equation~\ref{eqn:3Ddisp} the terms on the r.h.s. are related to rotation, self-gravity and internal pressure, respectively. The Toomre $Q$ parameter for the gas component can be defined as \citep{Toomre64a}:
\begin{equation}
Q_{\rm gas}\equiv\frac{\sigma_{\rm gas}\kappa}{\pi G \Sigma_{\rm gas}}\,.
\label{eqn:Q}
\end{equation}
In the thin disk approximation ($kh\ll1$), instability ($\omega^2 < 0$ in  \Eq{3Ddisp}) occurs on scales $k$ such that $Q < Q_{\rm crit}\simeq1$ (or equivalently $\omega^2 < 0$). Physically, this implies that the local gravitational field 
overcomes support from pressure and rotation, allowing fragmentation and collapse to happen. Similarly, for a collisionless fluid --- as an ensemble of stars --- the Toomre $Q$ parameter is defined by \citep{Toomre64a}:
\begin{equation}
Q_{\star} \equiv\frac{\sigma_{\star}\kappa}{3.36 G \Sigma_{\star}}\,.
\end{equation}

In our stability analysis, we account for the non-negligible disk thickness and for the combined effect of gas and stars; this is done adopting the approximation for an effective two-component $Q_{\rm eff}$ parameter \citep[i.e.,][see also \citealt{Inoue16a}]{Romeo11a}.
%
The effect of disk thickness modify the $Q$ parameter for gas/star by accounting for the vertical velocity dispersion:
\begin{equation}
T_{x} = \left\{
		\begin{array}{lccr}
			{\displaystyle 0.8 + 0.7\left(\frac{\sigma_{z}}{\sigma_{r}}\right)}      && & \mbox{if\ } \sigma_z \lesssim 0.5 \times \sigma_r \\ [1.25em]
			{\displaystyle 1 + 0.6\left(\frac{\sigma_{z}}{\sigma_{r}}\right)}        & & & \mbox{if\ } \sigma_z \gtrsim 0.5 \times \sigma_r
\\
		\end{array}
	\right.
\end{equation}
and
\begin{equation}
Q^{\rm thick}_{x} = T_{x} Q\,,
\end{equation}
with $x$ indicating either gas or stars.
% Note that w/ thickness, there's the a self-gravity dilution term in the dispersion relation, which in effect reduceds the surface gravity in the vertical direction. As such, it is easier to be stable against gravitaional instability. Thus, the criterion for instability $Q$ is reduced.
%
The combined effect of gas and stars can then be accounted for by writing
\begin{equation}\label{eqn:q_eff}
Q^{-1}_{\rm eff} =  \left\{
				\begin{array}{lccr}
					     {\displaystyle\frac{w}{Q^{\rm thick}_{\star}} + \frac{1}{Q^{\rm thick}_{\rm gas}}}      & & & \mbox{if\ }  Q^{\rm thick}_{\star} \geq Q^{\rm thick}_{\rm gas} \\ [0.75em]
                                               {\displaystyle\frac{1}{Q^{\rm thick}_{\star}} + \frac{w}{Q^{\rm thick}_{\rm gas}}}      & & & \mbox{if\ } Q^{\rm thick}_{\star} \leq Q^{\rm thick}_{\rm gas} \\
				\end{array}
			    \right.
\end{equation}
where the relative weight $w$ is defined as:
\begin{equation}
w\equiv\frac{2 \sigma_{\star} \sigma_{\rm gas}}{\sigma_{\star}^2 + \sigma_{\rm gas}^2},
\end{equation}
Conceptually, the thickness of a disk reduces the gravity in the vertical direction, thereby making it easier for a system to obtain stability, and thus lowering the criterion of the critical Toomre $Q_{\rm crit}$ from $\simeq$\,1 to 
0.67 \citep{Goldreich65a}.
%
On the other hand, including the contribution of the stellar component promotes gravitational instability, and thus, increasing the criterion of the Toomre $Q_{\rm crit}$, unless it is dynamically hot to ``compensate''. As a rule of thumb, $Q_{\rm crit}\eq1.34$ for $Q_{\rm gas}$\eq$Q_\star$.  % 0.67 * 2

>>>>

The onset of gravitational instability is 
%mm considered to be 
tightly connected to \SF \citep[e.g.,][]{Kennicutt89a, Wang94a, Li05b, Li06a}. 
%mm [rearranged to put derivation before result]
For axisymmetric modes, the dispersion relation for the growth of density perturbations in a rotating, turbulent disk of finite thickness $h$ is described by
\begin{equation}
\omega^2 = \kappa^2 - \frac{2\pi G \Sigma |k|}{1 + |k| h} + \sigma_{\rm disk}^2 k^2\,,
\label{eqn:3Ddisp}
\end{equation}
where $k$ is the wavenumber and $\kappa$ is the epicyclic frequency, defined as:
\begin{equation}
\kappa^2\equiv\frac{2\Omega}{\altomega}\frac{d}{d\altomega}\left(\altomega^2\Omega\right)\,.
\label{eqn:kappa}
\end{equation}
\citep{Romeo92a}.
\AF{$R$ is not the same as the MCC radius in the prev.\ Section} \MM{replaced $R$ with $\altomega$, and $\sigma$ with $\sigma_{\rm disk}$.}
In Equation~\ref{eqn:3Ddisp} the terms on the right hand side are related to rotation, self-gravity and internal pressure, respectively. 
%mm
    Heuristically, the instability can be understood by considering the scale
 at which gravitational potential overcomes the internal energy. Gravity dominates at scales $L > L_J$, where $L_J$ is the Jeans length. However, 
%mm centrifugal forces resulting from
   shearing by 
differential rotation in disk galaxies can stabilize perturbations 
%mm even on these scales. 
    that might otherwise collapse for $L > L_{\rm rot}$, where $L_{\rm rot}$ is set by $\kappa$.
%mm As a result, a disk is thought to be unstable on
    As a result, disks are unstable to gravitational collapse on 
scales between $L_J < L < L_{\rm rot}$.

From the dispersion relation, a parameter $Q$ can be derived such that $Q < 1$ when instability occurs, that reproduces this inequality to order unity.
For a collisionless fluid---such as an ensemble of stars--- this parameter is \citep{Toomre64a}
\begin{equation}
Q_{\star} \equiv\frac{\sigma_{\star}\kappa}{3.36 G \Sigma_{\star}}\,.
\end{equation}
The equivalent parameter for a collisional gas was derived by
%mm [corrected from Toomre64a to Goldreich65a (not Goldreich65b, which is about spiral arm formation)]
    \citep{Goldreich65a}:
\begin{equation}
Q_{\rm gas}\equiv\frac{\sigma_{\rm gas}\kappa}{\pi G \Sigma_{\rm gas}}\,.
\label{eqn:Q}
\end{equation}
In the thin disk approximation ($kh\ll1$), instability ($\omega^2 < 0$ in  \Eq{3Ddisp}) occurs on scales $k$ such that $Q < Q_{\rm crit}\simeq1$ (or equivalently $\omega^2 < 0$). 
%mm  In theoretical and observational works, gravitational instability has been widely discussed in the context of the Toomre $Q$ stability criterion \citep{Toomre64a, Goldreich65b} or its observable proxy
     A frequently used observable proxy for $Q$ is the ratio of disk circular velocity to root-mean-square velocity dispersion
$v_{\rm circ}/\sigma_{\rm disk}$
%mm , i.e., the circular velocity to the disk r.m.s. velocity dispersion \AF{This is not exactly the same $\sigma$ as before} \MM{see above} ratio 
\citep[e.g.,][]{GarciaBurillo03a, Genzel11a, Kassin12a, Leung19a}.
% https://ned.ipac.caltech.edu/level5/March15/Glazebrook/Glazebrook5.html
%
% from linear analysis

In our stability analysis, we account for the combined effect of gas and stars 
%mm 
     \citep[derived exactly by][]{Rafikov01}, 
and for the non-negligible disk thickness. This is done by adopting an approximation for an effective two-component $Q_{\rm eff}$ parameter \citep[i.e.,][see also \citealt{Inoue16a}]{Romeo11a}.
%
The effect of disk thickness modifies the $Q$ parameter for gas and stars by accounting for the vertical velocity dispersion:
\begin{equation}
T_{x} = \left\{
		\begin{array}{lccr}
			{\displaystyle 0.8 + 0.7\left(\frac{\sigma_{z}}{\sigma_{r}}\right)}      && & \mbox{if\ } \sigma_z \lesssim 0.5 \times \sigma_r \\ [1.25em]
			{\displaystyle 1 + 0.6\left(\frac{\sigma_{z}}{\sigma_{r}}\right)}        & & & \mbox{if\ } \sigma_z \gtrsim 0.5 \times \sigma_r
\\
		\end{array}
	\right.
\end{equation}
and
\begin{equation}
Q^{\rm thick}_{x} = T_{x} Q\,,
\end{equation}
with $x$ indicating either gas or stars.
% Note that w/ thickness, there's the a self-gravity dilution term in the dispersion relation, which in effect reduceds the surface gravity in the vertical direction. As such, it is easier to be stable against gravitaional instability. Thus, the criterion for instability $Q$ is reduced.
%
The combined effect of gas and stars can then be accounted for by writing
\begin{equation}\label{eqn:q_eff}
Q^{-1}_{\rm eff} =  \left\{
				\begin{array}{lccr}
					     {\displaystyle\frac{w}{Q^{\rm thick}_{\star}} + \frac{1}{Q^{\rm thick}_{\rm gas}}}      & & & \mbox{if\ }  Q^{\rm thick}_{\star} \geq Q^{\rm thick}_{\rm gas} \\ [0.75em]
                                               {\displaystyle\frac{1}{Q^{\rm thick}_{\star}} + \frac{w}{Q^{\rm thick}_{\rm gas}}}      & & & \mbox{if\ } Q^{\rm thick}_{\star} \leq Q^{\rm thick}_{\rm gas} \\
				\end{array}
			    \right.
\end{equation}
where the relative weight $w$ is defined as:
\begin{equation}
w\equiv\frac{2 \sigma_{\star} \sigma_{\rm gas}}{\sigma_{\star}^2 + \sigma_{\rm gas}^2},
\end{equation}
Conceptually, the finite disk thickness reduces the gravity in the vertical direction, thereby making it easier for a system to
%mm obtain
    maintain 
stability, and thus lowering the critical Toomre $Q_{\rm crit}$ from $\simeq$\,1 to 
0.67 \citep{Goldreich65a}.
%
On the other hand, including the contribution of the stellar component promotes gravitational instability, and thus, increases
%mm  the criterion of the Toomre
$Q_{\rm crit}$
%mm, unless it is dynamically hot to ``compensate''. 
As a rule of thumb, $Q_{\rm crit}\eq1.34$ for $Q_{\rm gas}$\eq$Q_\star$.  % 0.67 * 2
